\singlespacing A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}

### 2020-07-22 article.Rmd
Made some major revisions to the text

---
title: \singlespacing A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}
author: Henrik Kenneth Andersen
abstract: "This article provides an in-depth look at fixed effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed effects in the `lavaan` package for `R`. It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output. Online supplementary materials further discuss various common extentsions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/ random effects model. \\par  \\textbf{Keywords:} Fixed effects, structural equation modeling, lavaan, R, panel analysis" 
date: "`r Sys.Date()`"
output: 
  pdf_document:
    fig_caption: yes
    number_sections: yes
    includes: 
      in_header: header.tex
documentclass: article
classoption: a4paper
urlcolor: blue
link-citations: yes
bibliography: references.bib   
---

@@

Several years ago, @Curran2011 reflected positively on the growing use of panel studies in empirical social research. Some of the strengths of panel data are well-known, e.g., the ability to establish temporal precedence, increased statistical power and the reduction of potential alternative models. However, perhaps the greatest strength of panel data is that they allow for a more rigorous testing of substantive theories. Panel data, i.e., repeated measures of the same observed units (people, schools, firms, countries, etc.), allow researchers to decompose the error term into a part that stays constant within units and the part that changes over time. The part that does not change over time can be seen as the combined effect of all time-invariant influences (e.g., sex, date of birth, nationality) on the dependent variable. Fixed effects (FE) regression involves controlling for these time-invariant influences via a number of various methods. It thus drastically reduces the number of potential confounders of the relationship between variables of interest.

@@

There are a number of articles describing basic concept of panel model regression, and FE regression in SEM [e.g., @Allison2011; @Bollen2010; @Teachman2001]. This article is intended as a \textit{practical guide} for researchers looking for in-depth help with specifying FE models in SEM. It focuses on the `lavaan` [@R-lavaan] package for `R` [@R-base]. While `Mplus` [@Mplus] is arguably the most robust SEM software currently available (in terms of features like alignment, latent variable interactions, for example) , the `lavaan` package has many benefits. First, it and `R` are open source and completely free. For researchers dipping their toes into SEM, there is no financial barrier to try, and no risk if they decide it is not for them. Second, the implementation of `lavaan` in the larger `R` environment is an enormous advantage. Instead of poring over reams of plain text, copying out coefficients by hand, every part of the `lavaan` output is available as an object. This means that all aspects of the model, from fit indices, to coefficients and standard errors, to the model matrices, can be accessed and easily integrated into tables and plots. Furthermore, `R` can be used for a great deal of applications. It can be used to manage and manipulate as well as simulate data, perform symbolic algebra, run more traditional analyses (e.g., multiple regression, logistic regression, principal component analysis), etc. Once one is comfortable using `R`, there is no longer any need to switch between different software for data preparation and analysis.

The following article outlines the basic idea of FE regression, the particularities of FE in SEM, and shows its implementation in `lavaan`. Using [simulated data](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output. A number of potential extensions to the basic model, including relaxing various assumptions, dealing with measurement error in both the independent and dependent variables, as well as the inclusion of time-invariant predictors in the form of a hybrid fixed-/ random effects model, are shown in detail in the form of [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf).

@@

To begin, let us start by reviewing a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012] (we will return to this model in the [online section](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) of the article when we discuss loosening assumptions)
\begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
\end{align}
where $y_{it}$ is the dependent variable for unit $i, \ i = 1, ..., N$ at time $t, \ t = 1, ..., T$, $\bm{x}_{it}$ is a $1 \times K$ vector of time-varying covariates (which could include a constant) linked to the dependent variable by the $K \times 1$ vector of coefficients $\bm{\beta}$. $\bm{z}_{i}$ is a $1 \times M$ vector of time-invariant covariates linked to the dependent variable by the $M \times 1$ vector of coefficients in $\bm{\gamma}$, $\alpha_{i}$ represents the combined effect of all unobserved unit-specific variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.

@@

We typically make the following assumptions [see, e.g., @Wooldridge2002; @Schmidheiny2019] about this model:

- Linearity: the model is linear in its parameters and $\E[\varepsilon_{it}] = 0$ and $\E[\alpha_{i}] = 0$. 
- Independence: the observations are independent across individuals (assured by random sampling in the cross-section), but not necessarily across time. 
- Strict exogeneity: $\E[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] = 0$, i.e., the idiosyncratic error at each point in time is mean independent of $\bm{X}_{i}$ at \textit{all} points in time, $t = 1, ..., T$, as well as $\bm{z}_{i}$ and $\alpha_{i}$.  
- Error variance: homoscedastic and non-serially correlated errors, i.e., $\Var(\varepsilon_{i}|\bm{X}_{i},\bm{z}_{i}, \alpha_{i}) = \sigma^{2}_{\varepsilon} \bm{I}$.

@@

If we are willing to make some assumptions about the distribution of the individual effects and, more importantly, their lack of a relationship with the observed time-varying covariates, we are left with the random effects model [@Wooldridge2002; @Wooldridge2012; @Schmidheiny2019; @Skrondal2004]. Namely, we assume the individual effects are normally distributed, with a constant (homoscedastic) variance conditional on the model covariates, i.e., $\alpha_{i}|\bm{X}_{i}, \bm{z}_{i} \sim N(0, \sigma^{2}_{\alpha})$.^[As long as $\bm{X}_{i}$ contains a constant, the assumption that $\alpha_{i}$ has a mean of zero is unproblematic [@Wooldridge2012].] Further, we assume that the expectation of the individual effects conditional on the model covariates is zero, i.e., $\E[\alpha_{i}|\bm{X}_{i},\bm{z}_{i}] = 0$, which implies they are also uncorrelated.

The random effects model works by defining a composite error term: $\nu_{it} = \alpha_{i} + \varepsilon_{it}$ and rewriting the model in Equation \eqref{eq:gpm} as 
\begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \nu_{it}, \ \text{or} \\
\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\nu}_{i}.
\end{align}
By adding the individual effects to the composite error in each time period, the composite error becomes serially correlated over time, to give the following conditional covariance matrix for the errors, 
\begin{align}
\Var(\bm{\nu} | \bm{X}, \bm{z}) & = \bm{\Omega}_{\bm{\nu}} = 
\begin{pmatrix}
\bm{\Omega}_{\bm{\nu}_{1}} & \hdots & 0 & \hdots & 0 \\
\vdots & \ddots & \vdots & \ddots & \vdots \\
0 & \hdots & \bm{\Omega}_{\bm{\nu}_{i}} & \hdots & 0 \\
\vdots & \ddots & \vdots & \ddots & \vdots \\
0 & \hdots & 0 & \hdots & \bm{\Omega}_{\bm{\nu}_{N}}
\end{pmatrix}
\end{align}
with typical elements 
\begin{align}
\Var(\bm{\nu}_{i} | \bm{X}_{i}, \bm{z}_{i}) & = \bm{\Omega}_{\bm{\nu}_{i}} = 
\begin{pmatrix}
\sigma^{2}_{\nu} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\alpha} \\
\sigma^{2}_{\alpha} & \sigma^{2}_{\nu} & \hdots & \sigma^{2}_{\alpha} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma^{2}_{\alpha} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\nu}
\end{pmatrix} \label{eq:omeganui}
\end{align}
where $\sigma^{2}_{\nu} = \sigma^{2}_{\alpha} + \sigma^{2}_{\varepsilon}$. This means that in the conditional covariance matrix of the errors, given the time-varying and -invariant covariates, units over time will be correlated due to the individual effects. We should keep the covariance structure of the errors in mind as it will help make sense of the use of latent variables to decompose the dependent variable into between- and within-variance components, discussed below in Section \ref{fe-sem}.

@@

What if the individual effects are \textit{not} independent of the observed time-varying covariates, i.e., $\E[\alpha_{i}|\bm{X}_{i}] \ne 0$? Then grouping them in with the error will cause $\bm{\beta}$ to be biased [@Wooldridge2002; @Wooldridge2012]. In order to drop assumptions involving the individual effects, a number of methods are available (e.g., differencing, least squares dummy variable regression), but the most common approach is to \textit{demean} the equation [@Bruederl2015]. Demeaning involves subtracting the per-unit, over-time average from each of the model terms, i.e.,
\begin{align}
(y_{it} - \bar{y}_{i}) & = (\bm{x}_{it} - \bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \bar{\bm{z}}_{i})\bm{\gamma} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
\ddot{y}_{it} & = \bm{\ddot{x}}_{it}\bm{\beta} + \ddot{\varepsilon}_{it}, \ \text{or} \\
\bm{\ddot{y}}_{i} & = \bm{\ddot{X}}_{i}\bm{\beta} + \bm{\ddot{\varepsilon}}_{i} \label{eq:fe}
\end{align}
where, for example, $\bar{y}_{i} = T^{-1}\sum_{i = 1}^{T}y_{it}$ (and the other over-time means are calculated analogously), and the variables with the dots above them represent the demeaned versions. 
Because the average of something that does not change is that thing itself, the individual effects, along with any time-invariant predictors, get wiped out by the demeaning. This means that no assumptions about the relatedness of the model covariates and the unit-specific portion of the error are needed. The unbiasedness of the estimate is related solely to the strict exogeneity assumption imposed on the idiosyncratic errors, i.e., $\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}}_{it}] = \E[\ddot{\varepsilon}_{it}] = 0$ which also implies $\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Bruederl2015; @Wooldridge2002]. The downside to this approach is that no time-invariant predictors can be included in the model. In the [online section](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf), it will be discussed how to get around this restriction using SEM.

@@

We first need to convert the data from stacked, long-format vectors of length $NT$ into $T$ individual vectors of length $N$. To see why this is necessary, consider what effect this has on the vector of responses $y_{it}$. Let us, for a minute ignore any covariates and focus just on the dependent variable (a so-called 'intercept-only' model) so that we have $y_{it} = \alpha_{i} + \varepsilon_{it}$. When we convert the data to wide-format, we get $T$ individual equations,
\begin{align}
y_{1i} & = \alpha_{i} + \varepsilon_{1i} \\
y_{2i} & = \alpha_{i} + \varepsilon_{2i} \\
 & \vdots \\
y_{Ti} & = \alpha_{i} + \varepsilon_{Ti}. \label{eq:wide}
\end{align}
Because the idiosyncratic errors are assumed to be uncorrelated across units and across time, the covariance between any two of the new wide vectors $\Cov(y_{ti},y_{si}) = \Var(\alpha_{i}), \ t \ne s$. Otherwise, when $t = s$, the covariance $\Cov(y_{ti},y_{ti}) = \Var(\alpha_{i}) + \Var(\varepsilon_{ti})$. This is the structure we saw above in a typical element of $\bm{\Omega}_{\nu}$.

And in fact this is exactly how a latent variable is used to account for time-invariant unobserved heterogeneity. The dependent variable at each timepoint is regressed onto the latent variable, see Figure \ref{fig:fesem}. Here, the regression weights or 'factor loadings' are fixed to one to represent our assumption that the effect of the time-invariant unobserved heterogeneity is constant over time. It also means that the estimated variance of the latent variable is equal to the \textit{average covariance between the wide-format columns of the dependent variable over time}. If $y_{it} = \alpha_{i} + \varepsilon_{it}$ is the true data generating process, then the relationship between two units over time is just $\alpha$, regardless of the time distance. Referring back to the random effects structure of $\bm{\Omega}_{\bm{\nu}_{i}}$ in Equality \eqref{eq:omeganui} for a generic unit $i$, we see the covariance on all of the off-diagonals is $\sigma^{2}_{\alpha}$. And, as we know, the average of something that does not change is that thing itself. I.e., if $T(T-1)/2 = h$ is the number of elements on either the upper- or lower triangle of $\bm{\Omega}_{\bm{\nu}_{i}}$, then we have $h^{-1}\sum_{i=1}^{h}\sigma^{2}_{\alpha} = \frac{h \sigma^{2}_{\alpha}}{h} = \sigma^{2}_{\alpha}$.

>>>>

---
title: \singlespacing A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}
author: Henrik Kenneth Andersen
abstract: "This article provides an in-depth look at fixed effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed effects in the `lavaan` package for `R`. It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output. Online supplementary materials further discuss various common extentsions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/ random effects model. \\par  \\textbf{Keywords:} Fixed effects, structural equation modeling, lavaan, R, panel analysis" 
note: "Preprint version, please do not cite without permission."
date: "`r Sys.Date()`"
output: 
  pdf_document:
    fig_caption: yes
    number_sections: yes
    includes: 
      in_header: header.tex
documentclass: article
classoption: a4paper
urlcolor: blue
link-citations: yes
bibliography: references.bib   
---

@@

Several years ago, @Curran2011 reflected positively on the growing use of panel studies in empirical social research. Some of the strengths of panel data are well-known, e.g., the ability to establish temporal precedence, increased statistical power and the reduction of potential alternative models. However, perhaps the greatest strength of panel data is that they allow for a more rigorous testing of substantive theories. Panel data, i.e., repeated measures of the same observed units (people, schools, firms, countries, etc.), allow researchers to decompose the error term into a part that stays constant within units and the part that changes over time. The part that does not change over time can be seen as the combined effect of all time-invariant influences (e.g., sex, date of birth, nationality) on the dependent variable. Fixed effects (FE) regression involves controlling for these time-invariant influences via a number of various methods. It thus accounts for a likely and common source of bias.

@@

There are a number of articles describing basic concept of panel model regression, and FE regression in SEM [e.g., @Allison2011; @Bollen2010; @Teachman2001]. This article is intended as a \textit{practical guide} for researchers looking for in-depth help with specifying FE models in SEM. It focuses on the `lavaan` [@R-lavaan] package for `R` [@R-base]. While `Mplus` [@Mplus] is arguably the most robust SEM software currently available (in terms of features like alignment, latent variable interactions, for example) , the `lavaan` package has many benefits. First, like `R` it is open source and completely free. For researchers dipping their toes into SEM, there is no financial barrier to try, and no risk if they decide it is not for them. Second, the implementation of `lavaan` in the larger `R` environment is an enormous advantage. Instead of poring over reams of plain text, copying out coefficients by hand, every part of the `lavaan` output is available as an object. This means that all aspects of the model, from fit indices, to coefficients and standard errors, to the model matrices, can be accessed and easily integrated into tables and plots. Furthermore, `R` can be used for a great deal of applications. It can be used to manage and manipulate as well as simulate data, perform symbolic algebra, run more traditional analyses (e.g., multiple regression, logistic regression, principal component analysis), etc. Once one is comfortable using `R`, there is no longer any need to switch between different software for data preparation and analysis.

The following article outlines the basic idea of FE regression, the particularities of FE in SEM, and shows its implementation in `lavaan`. Using [simulated data](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output. One of the main strengths of panel SEM compared to the more traditional methods of panel analysis is its flexibility. Therefore, a number of potential extensions to the basic model, including relaxing various assumptions, dealing with measurement error in both the independent and dependent variables, as well as the inclusion of time-invariant predictors in the form of a hybrid fixed-/ random effects model, are shown in detail in the form of [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf).

@@

To begin, let us start by reviewing a general panel model [@Bollen2010], also referred to as the 'unobserved effects model' [@Wooldridge2012; @R-plm_a] (we will return to this model in the [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf) of the article when we discuss loosening assumptions)
\begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \alpha_{i} + \varepsilon_{it} \label{eq:gpm}
\end{align}
where $y_{it}$ is the dependent variable for unit $i, \ i = 1, ..., N$ at time $t, \ t = 1, ..., T$, $\bm{x}_{it}$ is a $1 \times K$ vector of time-varying covariates (which could include a constant) linked to the dependent variable by the $K \times 1$ vector of coefficients $\bm{\beta}$. $\bm{z}_{i}$ is a $1 \times M$ vector of time-invariant covariates linked to the dependent variable by the $M \times 1$ vector of coefficients in $\bm{\gamma}$, $\alpha_{i}$ represents the combined effect of all unobserved time-constant variables affecting the dependent variable and $\varepsilon_{it}$ is the idiosyncratic error.

@@

Consistency of the following models requires the assumption of strict exogeneity, although what constitutes strict exogeneity differs between the random and fixed effects setups. Each assumption will be discussed shortly. Apart from that, we typically make the following assumptions [see, e.g., @Wooldridge2002; @Schmidheiny2019] about this model:

- Linearity: the model is linear in its parameters. 
- Independence: the observations are independent across individuals (assured by random sampling in the cross-section), but not necessarily across time. 
- The usual rank condition: we have more observations than independent variables and there is no perfect collinearity between any of the independent variables.

@@

For the random effects (RE) model, we define a composite error term: $\nu_{it} = \alpha_{i} + \varepsilon_{it}$ and rewrite the model in Equation \eqref{eq:gpm} as 
\begin{align}
y_{it} & = \bm{x}_{it}\bm{\beta} + \bm{z}_{i}\bm{\gamma} + \nu_{it}, \ \text{or} \\
\bm{y}_{i} & = \bm{X}_{i}\bm{\beta} + \bm{Z}_{i}\bm{\gamma} + \bm{\nu}_{i},
\end{align}
where $\bm{\nu}_{i} = \alpha_{i} \bm{\iota}_{T} + \bm{\varepsilon}_{i}$ and $\bm{\iota}_{T}$ is a $T \times 1$ vector of ones. The strict exogeneity assumption in the RE model implies
\begin{align}
\E[\varepsilon_{it} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = 0, \\
\E[\alpha_{i} |\bm{X}_{i}, \bm{z}_{i}] = \E[\alpha_{i}] & = 0,
\end{align}
where $\bm{X}_{i} = \bm{x}_{i1}, ..., \bm{x}_{iT}$. For both parts, the assumption that the unconditional expectations are 0 is unproblematic as long as a constant is included in the regression. The first part says the idiosyncratic errors at each timepoint are assumed to be independent of the explanatory variables at \textit{all} timepoints which is stronger than just assuming that they are \textit{contemporaneously} independent. This implies that they are also uncorrelated, i.e., $\E[\bm{x}_{is}^{\intercal}\varepsilon_{it}] = \bm{0}$ and $\E[\bm{z}_{i}^{\intercal}\varepsilon_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Wooldridge2002; @Bruederl2015]. We assume the idiosyncratic errors are further independent of the individual effects, which implies $\E[\alpha_{i}\varepsilon_{it}] = 0$.

The second part is the potentially controversial assumption: it states that the individual effects are uncorrelated with the independent variables. We can use an example to show why this is often controversial: If we are interested in the question of whether married men earn more than unmarried men, then the second part of the strict exogeneity assumption means that a man's marriage status would have to be uncorrelated with all the time-invariant characteristics that could potentially make that man an attractive marriage candidate in the first place; e.g., looks, personality, family's status, profession, etc. [@Bruederl2015].^[Assuming, for the sake of argument, that these characteristics are constant over time.]

From what we have discussed so far, the $T \times T$ covariance matrix of the errors $\bm{\Omega}_{i} = \E[\bm{\nu}_{i}\bm{\nu}_{i}^{\intercal}]$ can be constructed. However, the standard random effects model adds the additional assumptions 
\begin{alignat}{3}
\E[\varepsilon_{it}^{2} | \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}^{2}] && = \sigma_{\varepsilon}^{2}, && \ t = 1, ..., T, \\
\E[\varepsilon_{it}\varepsilon_{is}| \bm{X}_{i}, \bm{z}_{i}, \alpha_{i}] & = \E[\varepsilon_{it}\varepsilon_{is}] && = 0, && \ \forall \ t \ne s 
\end{alignat}
i.e., the idiosyncratic errors are conditionally homoscedastic and serially uncorrelated and 
\begin{alignat}{2}
\E[\alpha_{i}^{2}| \bm{X}_{i},\bm{z}_{i}] & = \E[\alpha_{i}^{2}] && = \sigma^{2}_{\alpha}
\end{alignat}
i.e., the individual effects are conditionally homoscedastic (they are necessarily serially correlated as long as $\sigma_{\alpha}^{2} > 0$) we arrive at the typical random effects structure of the $NT \times NT$ matrix $\bm{\Omega}$:

\begin{align}
\bm{\Omega} & = 
\begin{pmatrix}
\bm{\Omega}_{1} & \hdots & 0 & \hdots & 0 \\
\vdots & \ddots & \vdots & \ddots & \vdots \\
0 & \hdots & \bm{\Omega}_{i} & \hdots & 0 \\
\vdots & \ddots & \vdots & \ddots & \vdots \\
0 & \hdots & 0 & \hdots & \bm{\Omega}_{N}
\end{pmatrix}
\end{align}
with $T \times T$ typical elements 
\begin{align}
\bm{\Omega}_{i} & = 
\begin{pmatrix}
\sigma^{2}_{\nu} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\alpha} \\
\sigma^{2}_{\alpha} & \sigma^{2}_{\nu} & \hdots & \sigma^{2}_{\alpha} \\
\vdots & \vdots & \ddots & \vdots \\
\sigma^{2}_{\alpha} & \sigma^{2}_{\alpha} & \hdots & \sigma^{2}_{\nu}
\end{pmatrix} \label{eq:omeganui}
\end{align}
where $\sigma^{2}_{\nu} = \sigma^{2}_{\alpha} + \sigma^{2}_{\varepsilon}$ [@Wooldridge2002; @Schmidheiny2019]. This means that in the conditional covariance matrix of the errors, given the time-varying and -invariant covariates, units over time will be correlated due to the individual effects. We should keep the covariance structure of the errors in mind as it will help make sense of the use of latent variables to decompose the dependent variable into between- and within-variance components, discussed below in Section \ref{fe-sem}.

Estimation of the RE model can be done using feasible generalized least squares (GLS) in which the two unknowns in $\bm{\Omega}$, $\sigma_{\alpha}^{2}$ and $\sigma_{\nu}^{2}$, are first estimated using pooled ordinary least squares (pooled OLS or POLS), where^[Normally a degrees-of-freedom correction is applied by subtracting off the number of independent variables that is negligible in large N samples. It is ignored here for the sake of simplicity.]
\begin{align}
\hat{\sigma}_{\nu}^{2} & = \frac{1}{NT}\sum_{i = 1}^{N}\sum_{t = 1}^{T}\hat{\nu}_{it}^{2}, \\
\hat{\sigma}_{\varepsilon}^{2} & = \frac{1}{NT - N}\sum_{i = 1}^{N}\sum_{t = 1}^{T}(\hat{\nu}_{it} - \bar{\hat{\nu}}_{i})^{2}
\end{align}
and $\hat{\nu}_{it} = y_{it} - \bm{x}_{it}\hat{\bm{\beta}}_{\tiny{POLS}} - \bm{z}_{i}\hat{\bm{\gamma}}_{\tiny{POLS}}$, $\bar{\hat{\nu}}_{i} = T^{-1}\sum_{t = 1}^{T}\hat{\nu}_{it}$ and $\hat{\sigma}^{2}_{\alpha} = \hat{\sigma}^{2}_{\nu} - \hat{\sigma}^{2}_{\varepsilon}$. Then, the coefficients are estimated using those estimates in the variance matrix $\hat{\bm{\Omega}}$:
\begin{align}
\begin{pmatrix}
\hat{\bm{\beta}}_{RE} \\
\hat{\bm{\gamma}}_{RE}
\end{pmatrix} & = 
(\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{W})^{-1}\bm{W}^{\intercal}\hat{\bm{\Omega}}^{-1}\bm{y}
\end{align}
where $\bm{W} = \begin{pmatrix}\bm{X} & \bm{Z}\end{pmatrix}$ and $\bm{X}$ is $NT \times K$ and $\bm{Z}$ is $NT \times M$ and $\bm{y}$ is $NT \times 1$.

In practice, however, computational problems can arise with large cross-sectional samples, where it can become difficult to invert the $\hat{\bm{\Omega}}$ matrix. One solution is to use 'partial-demeaning' to transform the data before performing simple OLS:
\begin{align}
(y_{it} - \theta\bar{y}_{i}) & = (\bm{x}_{it} - \theta\bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \theta \bar{\bm{z}}_{i})\bm{\gamma} + (\varepsilon_{it} - \theta \bar{\varepsilon}_{i}) \label{eq:partialdemean}
\end{align} 
where $\theta = 1 - [\sigma_{\alpha}^{2}/(\sigma_{\alpha}^{2} + T \sigma_{\varepsilon}^{2})]^{1/2}$, and $\bar{y}_{i} = T^{-1}\sum_{t = 1}^{T}y_{it}$, $\bar{\bm{x}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{x}_{it}$, $\bar{\bm{z}}_{i} = T^{-1}\sum_{t = 1}^{T}\bm{z}_{i}$ and $\bar{\varepsilon}_{i} = T^{-1}\sum_{t = 1}^{T}\varepsilon_{it}$ [@R-plm_a].

The RE model can also be estimated in the maximum likelihood framework, where in the associated literature panel models are generally referred to as either mixed models, hierarchical models or longitudinal models. The typical RE model discussed here is the equivalent to a mixed model with random intercepts and fixed slopes [@R-plm_a]. Under the assumption of normality, along with homoscedasticity and serially uncorrelated errors, the maximum likelihood estimator is the same as the OLS estimator. For more on the topic of mixed models, see for example @R-lme4, @Bates2010.

@@

For fixed effects, we assume that the individual effects are \textit{not independent} of the model covariates, i.e., $\E[\alpha_{i}|\bm{X}_{i},\bm{z}_{i}] \ne \E[\alpha_{i}] \ne 0$. Under this assumption, grouping the individual effects in with the composite error will cause the coefficients of interest, here specifically $\bm{\beta}$ to be inconsistent [@Wooldridge2002; @Wooldridge2012]. We write the model therefore again in terms of the general panel model in Equation \eqref{eq:gpm} with separate individual effects and idiosyncratic error terms. In order to drop assumptions involving the individual effects, a number of methods are available (e.g., differencing, least squares dummy variable regression), but the most common approach is to \textit{demean} the equation [@Bruederl2015]. Demeaning is the same as the transformation applied in Equation \eqref{eq:partialdemean} in the special case where $\theta = 1$. I.e., demeaning involves subtracting the per-unit, over-time average from each of the model terms, i.e.,
\begin{align}
(y_{it} - \bar{y}_{i}) & = (\bm{x}_{it} - \bar{\bm{x}}_{i})\bm{\beta} + (\bm{z}_{i} - \bar{\bm{z}}_{i})\bm{\gamma} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
\ddot{y}_{it} & = \bm{\ddot{x}}_{it}\bm{\beta} + \ddot{\varepsilon}_{it}, \ \text{or} \\
\bm{\ddot{y}}_{i} & = \bm{\ddot{X}}_{i}\bm{\beta} + \bm{\ddot{\varepsilon}}_{i} \label{eq:fe}
\end{align}
where the over-time averages are calculated the same as above, and the variables with the dots above them represent the demeaned versions. 
Because the average of something that does not change is that thing itself, the individual effects, along with any time-invariant predictors, get wiped out by the demeaning. This means that no assumptions about the relatedness of the model covariates and the unit-specific portion of the error are needed. Consistency of the estimates is related solely to the strict exogeneity assumption imposed on the idiosyncratic errors, i.e., $\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}}_{it}] = \E[\ddot{\varepsilon}_{it}] = 0$ which also implies $\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Bruederl2015; @Wooldridge2002].

Having demeaned the data, the typical FE estimator is POLS on the transformed data
\begin{align}
\bm{\beta}_{FE} & = (\ddot{\bm{X}}^{\intercal}\ddot{\bm{X}})^{-1}\ddot{\bm{X}}^{\intercal}\ddot{\bm{y}}
\end{align}
[@Bruederl2015]. The downside to this approach is that no time-invariant predictors can be included in the model. However, there are alternative approaches in the random effects and mixed model frameworks that allow them to be included. These models are sometimes referred to as 'within-between' or 'hybrid' models, often based on the Chamberlain -@Chamberlain1980 and Mundlak -@Mundlak1978 approaches, see for example @Bell2018; @Allison2011; @Schunck2013; @Enders2007. In the [online appendix](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf), it will be discussed how to also get around this restriction using SEM.

@@

We first need to convert the data from stacked, long-format vectors of length $NT$ into $T$ individual vectors of length $N$. To see why this is necessary, consider what effect this has on the vector of responses $y_{it}$. Let us, for a minute ignore any covariates and focus just on the dependent variable (a so-called 'intercept-only' or 'null' model) so that we have $y_{it} = \alpha_{i} + \varepsilon_{it}$. When we convert the data to wide-format, we get $T$ individual equations,
\begin{align}
\bm{y}_{t} & = \bm{\alpha} + \bm{\varepsilon}_{t} \\
\begin{bmatrix}
y_{1t} \\
y_{2t} \\
\vdots \\
y_{Nt}\end{bmatrix} & = 
\begin{bmatrix}
\alpha_{1} \\
\alpha_{2} \\
\vdots \\
\alpha_{N}
\end{bmatrix} + 
\begin{bmatrix}
\varepsilon_{1t} \\
\varepsilon_{2t} \\
\vdots \\
\varepsilon_{Nt}
\end{bmatrix} \label{eq:wide}
\end{align}
for each $t = 1, 2, ..., T$. Because the idiosyncratic errors are assumed to be uncorrelated across units and across time, the covariance between any two of the new wide vectors $\Cov(y_{ti},y_{si}) = \Var(\alpha_{i}), \ t \ne s$. Otherwise, when $t = s$, the covariance $\Cov(y_{ti},y_{ti}) = \Var(\alpha_{i}) + \Var(\varepsilon_{ti})$. This is the structure we saw above in a typical element of $\bm{\Omega}$.

And in fact this is exactly how a latent variable is used to account for time-invariant unobserved heterogeneity. The dependent variable at each timepoint is regressed onto the latent variable, see Figure \ref{fig:fesem}. Here, the regression weights or 'factor loadings' are fixed to one to represent our assumption that the effect of the time-invariant unobserved heterogeneity is constant over time. It also means that the estimated variance of the latent variable is equal to the \textit{average covariance between the wide-format columns of the dependent variable over time}. If $y_{it} = \alpha_{i} + \varepsilon_{it}$ is the true data generating process, then the relationship between two units over time is just $\alpha$, regardless of the time distance. Referring back to the random effects structure of $\bm{\Omega}_{i}$ in Equality \eqref{eq:omeganui} for a generic unit $i$, we see the covariance on all of the off-diagonals is $\sigma^{2}_{\alpha}$. And, as we know, the average of something that does not change is that thing itself. I.e., if $T(T-1)/2 = h$ is the number of elements on either the upper- or lower triangle of $\bm{\Omega}_{i}$, then we have $h^{-1}\sum_{i=1}^{h}\sigma^{2}_{\alpha} = \frac{h \sigma^{2}_{\alpha}}{h} = \sigma^{2}_{\alpha}$.
